---
title: "Grad Project"
author: "Emily Lines"
date: "4/20/2020"
output: html_document
---

```{r load data, warning = FALSE, message = FALSE}

library(tidyverse)
ncaa_old <- read_csv("cbb.csv")

```

I've imported a dataset with stats from the 2015, 2016, 2017, 2018, and 2019 NCAA Men's DI basketball seasons. I want to find out if we can predict whether or not a team will make the NCAA post-season tournament from the other variables in the dataset. 

```{r update data}

ncaa <- ncaa_old %>%
  mutate(tournament = ifelse(ncaa_old$POSTSEASON %in% 
                  c('2ND', 'Champions', 'E8', 'F4', 'R64', 'R32', 'S16'), 1, 0))
ncaa$POSTSEASON = NULL
ncaa$SEED = NULL

```

I updated the dataset to create a column of dummy variables for whether or not a team made the tournament. Previously, `POSTSEASON` had data on which round the team was eliminated in the tournament, but I am only concerned about if they made the tournament. I then deleted the old `POSTSEASON` column. I also deleted `SEED` (a team's ranking in the tournament) because again I am only concerned about whether or not they made the tournament. 













